---
title: "ChannelGeometry"
format: gfm
---

This document provides a quick look at some of the results reported in Lex's Sprague_Reach_Habitat_updated.csv file. Let's start with the ODFW reach data.

```{r}
library(data.table)
library(stats)
library(ggplot2)
library(patchwork)

datafile <- "/Users/alexahaucke/Documents/GitHub/ChannelGeometry/Sprague_Reach_Habitat_updated.csv"
data <- as.data.table(read.csv(datafile))
reach_width <- data[ACW>0, .(AREA_SQKM, SPRAG14RCH, ACW)]
reach_widths <- reach_width[, .(width = mean(ACW, na.rm = TRUE)), by = SPRAG14RCH]
reach_areas <- reach_width[, .(area = mean(AREA_SQKM, na.rm = TRUE)), by = SPRAG14RCH]
width_reach <- merge(reach_widths, areas, by = "SPRAG14RCH")
p_pnts <- ggplot(width_reach, aes(x=area,y=width)) +
  geom_point(shape=21, size=3, color='black', fill='gray') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_box <- ggplot(width_reach, aes(y=width)) +
  geom_boxplot(outlier.color="black", outlier.shape=8, outlier.size=3) +
  labs(title="ODFW Reach Width") +
  theme(axis.title.y = element_blank())
p_pnts + p_box
```
The 17-m wide reach appears to be an outlier compared to the other values. 
```{r}
# Remove the apparent outlier
width_reach <- width_reach[width < 17]

# Fit a power function to the data, starting values from Excel
m_width <- nls(width ~ a * area^b,
               data = width_reach,
               start = list(a = 4.4, b = 0.0354))
m_width

# Plot the data and the fitted line
fit <- predict(m_width, newdata=width_reach)
p_width_reach <- ggplot(width_reach, aes(x=area,y=width)) +
  geom_point(shape=21, size=3, color='black', fill='gray') +
  geom_line(aes(x=width_reach$area, y=fit), linewidth=1.3, color='black') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_width_reach
```



##Width Regressions
Let's compare these results to other survey data.

```{r}
# Stack the widths
width_long <- melt(data, 
                   id.vars = c("SPRAG14RCH", "AREA_SQKM"),
                   measure.vars = c("ACW", "AC_WIDTH", "MaskWidth"),
                   variable.name = "width_type",
                   value.name = "width")
width_long <- width_long[width > 0]

# Then average by reach
width_long <- width_long[, .(width = mean(width, na.rm = TRUE),
                              area = mean(AREA_SQKM, na.rm = TRUE)), by = SPRAG14RCH]

```

Rerun the model

```{r}
# (Optionally) remove outliers
width_long <- width_long[width < 17]

# Fit power model again
m_width <- nls(width ~ a * area^b,
               data = width_long,
               start = list(a = 4.4, b = 0.0354))
summary(m_width)
```


```{r}
##Create width model
synthetic_width <- function(DA) {
  synthetic_widths <- 4.044*DA^0.08209
  return(synthetic_widths)
}

synthetic_widths <- synthetic_width(areas$area)

```
Castro et al. (2001) created regression models based on contributing areas, as well as width and heights. To compare our model to their model, let's run their model with our contributing area data.

```{r}
# Use equations for western interior basin and ranges, base on figure 4 in Castro & Jackson

# C&J report discharge in cubic feet per second and drainage area in square miles.
# Their Table 4: Q = 13.05*(DA^0.77). We want cubic meters per second 
# as a function of square kilometers. There are 0.0283168 cubic meters per cubic foot.
# There are 0.386102 square miles per square kilometer.
castro_flow <- function(DA) {
  castro_flows = 0.0283168*13.05*((0.386102*DA)^0.77)
  return(castro_flows)
}  

# C&J Table 4: w = 3.27*(DA^0.51), with width (w) in feet and 
# drainage area (DA) in square miles. There are 0.3048 meters per foot.
# To get width in meters: w(m) = (m/ft)*3.27*((sq miles/sq km)*DA(sq km))^0.51
castro_width <- function(DA) {
  castro_widths = 0.3048*3.27*((0.386102*DA)^0.51)
  return(castro_widths)
}  

#C&J Table 4: d = 0.79*(DA^0.24). d in feet, convert to meters
castro_depth <- function(DA) {
  castro_depths = 0.3048*0.79*((0.386102*DA)^0.24)
  return(castro_depths)
}  

castro_widths <- castro_width(areas$area)
castro_depths <- castro_depth(areas$area)

```

To determine the differences between our values and those generated by this model, let's just take a look at the value distributions.

```{r}
hist(castro_widths)

hist(synthetic_widths)

hist(width_long$width)
```
Based on these charts, we can see that the Castro widths are mostly above 20 m, which doesn't really make sense based on the surveyed widths. There is a chance that I messed up the unit conversions with Castro, but nothing immediately springs to mind.

Next, let's see if the different surveys give different model results. First, the reach survey:
```{r}
# Fit power model again
m_width_reach <- nls(width ~ a * area^b,
               data = width_reach,
               start = list(a = 4.4, b = 0.0354))
summary(m_width_reach)
```



```{r}
hab_width <- data[AC_WIDTH>0, .(AREA_SQKM, SPRAG14HAB, AC_WIDTH)]
hab_widths <- hab_width[, .(width = mean(AC_WIDTH, na.rm = TRUE)), by = SPRAG14HAB]
hab_areas <- hab_width[, .(area = mean(AREA_SQKM, na.rm = TRUE)), by = SPRAG14HAB]
width_habitat <- merge(hab_widths, hab_areas, by = "SPRAG14HAB")
# (Optionally) remove outliers
width_habitat <- width_habitat[width < 17]

# Fit power model again
m_width_habitat <- nls(width ~ a * area^b,
               data = width_habitat,
               start = list(a = 4.4, b = 0.0354))
summary(m_width_habitat)
```


```{r}
mask_widths <- data[MaskWidth>0, .(AREA_SQKM, SPRAG14RCH, MaskWidth)]
mask_widths <- data[, .(width = mean(MaskWidth, na.rm = TRUE)), by = SPRAG14RCH]
width_mask <- merge(mask_widths, reach_areas, by = "SPRAG14RCH")

# (Optionally) remove outliers
width_mask <- width_mask[width < 17]
width_mask <- width_mask[width > 0]

# Fit power model again
m_width_mask <- nls(width ~ a * area^b,
               data = width_mask,
               start = list(a = 4.4, b = 0.0354))
summary(m_width_mask)
```
All of the models give coefficients between 4.2 and 4.6, and the powers are between 0.03 and 0.05. This suggests that our models are roughly consistent with each other. However, the reach model has the lowest coefficient and the highest power. Maybe this could be caused by different survey methods?

We can compare the predictions for these models to the data that they have been trained on.
```{r}
# Plot the data and the fitted line
fit_habitat <- predict(m_width_habitat, newdata=width_habitat)
fit_reach <- predict(m_width_reach, newdata=width_reach)
fit_mask <- predict(m_width_mask, newdata=width_mask)

# Add predicted values to the same data frame
width_reach[, fit := predict(m_width_reach, newdata = width_reach)]
width_habitat[, fit := predict(m_width_habitat, newdata = width_habitat)]
width_mask[, fit := predict(m_width_mask, newdata = width_mask)]

p_width_reach <- ggplot() +
  geom_point(aes(x=width_reach$area, y=width_reach$width),shape=21, size=3, color='black', fill='gray') +
  geom_line(aes(x=width_reach$area, y=width_reach$fit), linewidth=1.3, color='black') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_width_reach
  
p_width_habitat <- ggplot() +
  geom_point(aes(x=width_habitat$area, y=width_habitat$width),shape=21, size=3, color='red', fill='pink') +
  geom_line(aes(x=width_habitat$area, y=width_habitat$fit), linewidth=1.3, color='red') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_width_habitat
  
p_width_mask <- ggplot() +
  geom_point(aes(x=width_mask$area, y=width_mask$width),shape=21, size=3, color='blue', fill='lightblue') +
  geom_line(aes(x=width_mask$area, y=width_mask$fit), linewidth=1.3, color='blue') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_width_mask
```
This shows the data and the regressions created based on them. 


##Depth Regressions 
Next, let's try doing the same with depths.

```{r}
# Take a look at the channel depth values
reach_depth <- data[ACH>0, .(AREA_SQKM, SPRAG14RCH, ACH)]
reach_depths <- reach_depth[, .(depth = mean(ACH, na.rm = TRUE)), by = SPRAG14RCH]
reach_areas <- reach_depth[, .(area = mean(AREA_SQKM, na.rm = TRUE)), by = SPRAG14RCH]
depth_reach <- merge(reach_depths, reach_areas, by = "SPRAG14RCH")
p_pnts <- ggplot(depth_reach, aes(x=area,y=depth)) +
  geom_point(shape=21, size=3, color='black', fill='gray') +
  labs(title="ODFW Reach Depth vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Depth (m)")
p_pnts
```

Hmm, this shows channel depth decreasing with increasing contributing area. Perhaps they weren't measuring bank-full depth, but including channel incision in their measurements. 

Let's compare this to the habitat surveys
```{r}
# Take a look at the channel depth values
hab_depth <- data[AC_HEIGHT>0, .(AREA_SQKM, SPRAG14HAB, AC_HEIGHT)]
hab_depths <- hab_depth[, .(depth = mean(AC_HEIGHT, na.rm = TRUE)), by = SPRAG14HAB]
hab_areas <- hab_depth[, .(area = mean(AREA_SQKM, na.rm = TRUE)), by = SPRAG14HAB]
depth_hab <- merge(hab_depths, hab_areas, by = "SPRAG14HAB")
p_pnts2 <- ggplot(depth_hab, aes(x=area,y=depth)) +
  geom_point(shape=21, size=3, color='black', fill='gray') +
  labs(title="ODFW Reach Depth vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Depth (m)")
p_pnts2
```

The results look pretty similar for the habitat and reach surveys.

```{r}
# Fit power model for both
m_depth_reach <- nls(depth ~ a * area^b,
               data = depth_reach,
               start = list(a = 4.4, b = 0.0354))
summary(m_depth_reach)

m_depth_hab <- nls(depth ~ a * area^b,
               data = depth_hab,
               start = list(a = 4.4, b = 0.0354))
summary(m_depth_hab)


# Add predicted values to the same data frame
depth_reach[, fit := predict(m_depth_reach, newdata = depth_reach)]
depth_hab[, fit := predict(m_depth_hab, newdata = depth_hab)]

p_depth_reach <- ggplot() +
  geom_point(aes(x=width_reach$area, y=width_reach$width),shape=21, size=3, color='black', fill='gray') +
  geom_line(aes(x=width_reach$area, y=width_reach$fit), linewidth=1.3, color='black') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_depth_reach
  
p_depth_habitat <- ggplot() +
  geom_point(aes(x=width_habitat$area, y=width_habitat$width),shape=21, size=3, color='red', fill='pink') +
  geom_line(aes(x=width_habitat$area, y=width_habitat$fit), linewidth=1.3, color='red') +
  labs(title="ODFW Reach Width vs Contributing Area",
       x="Contributing Area (sq km)",
       y="Mean Reach Width (m)")
p_depth_habitat



```
The models generally show similar results.
